Family vs Phylum Models
=======================

## Setup

### Load Packages

```{r}
library(tidyverse)
library(glmmTMB)
library(emmeans)
library(AfterSl1p)
source('scripts/stats/functions.R')
theme_set(theme_bw())
```

## Read in data

### Sample data

```{r}
# sample data
dat = read.delim('data/clean/coloniz_model.txt')
dat_map = (dat
           %>% select(Sample, Group, Category, Cage)
           %>% unique())
rm(dat)
```

### Bins/MAGs

```{r}
bin_tax = read.delim('data/clean/binQT.txt')

bin_cov = read.delim('data/clean/binCover.txt')
bin_cov = filter(bin_cov, Sample %in% dat_map$Sample)
all_mags = read.delim('data/clean/MAGids.txt', header = FALSE)
colnames(all_mags) = 'BinLab'
mag_cov = read.delim('data/clean/MAGCover.txt')
mag_cov = filter(mag_cov, Sample %in% dat_map$Sample)
```

### Species Taxonomy

```{r}
spc_dat = read.delim('data/clean/speciesAbund.txt')
spc_dat = (spc_dat
           %>% mutate(feature = gsub('[a-z]__','',feature))
           %>% separate_wider_delim(feature, '|', 
                                    names = c('Kingdom', 'Phylum','Class', 
                                              'Order', 'Family', 'Genus', 
                                              'Species', 'Strain'), 
                                    too_few = 'align_start')
           %>% mutate(Sample = case_when(Sample == 'Wild116' ~ 'DonorB',
                                         Sample == 'X608' ~ 'DonorA',
                                         TRUE ~ Sample)))
spc_tax = (spc_dat
           %>% select(Kingdom:Strain)
           %>% unique())
```

## Wrangle Taxonomy

### Bins

```{r}
### Check if families are unique
chk_fams = (bin_tax
            %>% select(Phylum:Family)
            %>% unique())
dim(chk_fams)
n_distinct(chk_fams$Family)

### They are. I don't need to carry all the taxon info.
bin_tax = select(bin_tax, BinLab, Family, Phylum)
bin_fam_phy = (bin_tax
           %>% select(Family, Phylum)
           %>% unique())

### filter down to phyla that contain more than one family. Otherwise it's just
### the same data twice.

bin_fam_phy = (bin_fam_phy
               %>% count(Phylum)
               %>% filter(n > 1)
               %>% left_join(bin_fam_phy))
```

### Species

```{r}
### Check if families are unique
chk_fams = (spc_tax
            %>% select(Phylum, Family)
            %>% unique())
dim(chk_fams)
n_distinct(chk_fams$Family)
```

There are two sets of families that have the same names but different phyla. Let's see:

```{r}
chk_fams %>% count(Family) %>% filter(n > 1) %>% left_join(chk_fams)
```

I don't really understand how something can have a family assignment but lack a
phylum assignment, but what I'm going to do is filter out everything that lacks
a phylum or family assignment.

```{r}
spc_tax = (spc_tax
           %>% filter(!grepl('unclassified', Phylum),
                      !grepl('unclassified', Family)))
```

## Create database counts and find common families and phyla

### Bins

Find families and phyla that are present in both donors:

```{r}
head(bin_tax)
bin_don = (bin_tax
       %>% mutate(Donor = sub('Don', 'Donor', (substr(BinLab, 1, 4)))))

bin_fam_both = find_don_shared(bin_don, 'Family')
bin_phy_both = find_don_shared(bin_don, 'Phylum')
```

Make sure the phyla have at least 2 families:

```{r}
bin_fam_both = bin_fam_both[bin_fam_both %in% bin_fam_phy$Family]
bin_phy_both = bin_phy_both[bin_phy_both %in% bin_fam_phy$Phylum]
```

Get db sizes for the two donors

```{r}
bin_fam_size = (bin_don
             %>% filter(Family %in% bin_fam_both)
             %>% count(Donor, Family, name = 'size'))
bin_phy_size = (bin_don
             %>% filter(Phylum %in% bin_phy_both)
             %>% count(Donor, Phylum, name = 'size'))
```

### MAGs

```{r}
mag_tax = (bin_tax
           %>% filter(BinLab %in% all_mags$BinLab))
mag_don = (bin_don
           %>% filter(BinLab %in% all_mags$BinLab))

mag_fam_both = find_don_shared(mag_don, 'Family')
mag_fam_both = mag_fam_both[mag_fam_both %in% bin_fam_phy$Family]
mag_phy_both = find_don_shared(mag_don, 'Phylum')
mag_phy_both = mag_phy_both[mag_phy_both %in% bin_fam_phy$Phylum]
mag_fam_size = (mag_don
             %>% filter(Family %in% mag_fam_both)
             %>% count(Donor, Family, name = 'size'))
mag_phy_size = (mag_don
             %>% filter(Phylum %in% mag_phy_both)
             %>% count(Donor, Phylum, name = 'size'))
```

### Species

Start by splitting the data into donor and recipient and only keep those that
meet the cutoff

```{r}
head(spc_dat)
spc_don = (spc_dat
           %>% filter(startsWith(Sample, 'Don'),
                      coverage > 0.1)
           %>% rename(Donor = Sample))
spc_rec = (spc_dat
           %>% filter(!startsWith(Sample, 'Don'),
                      coverage > 0.1)
           %>% filter(Sample %in% dat_map$Sample))
```

Identify strains that are common to the two donors for elimination

```{r}
spc_don = (spc_don
           %>% select(-Donor, Strain)
           %>% unique()
           %>% count(Strain)
           %>% filter(n == 1)
           %>% left_join(spc_don)
           %>% select(Strain, Family, Phylum, Donor))
```

Remove unclassified 

```{r}
spc_don = (spc_don
           %>% filter(!grepl('unclassified', Phylum),
                      !grepl('unclassified', Family)))
```


Limit ourselves to phyla with more than one family

```{r}
spc_fam_phy = (spc_don
               %>% select(Family, Phylum)
               %>% unique())
spc_fam_phy = (spc_fam_phy
               %>% count(Phylum)
               %>% filter(n > 1)
               %>% left_join(spc_fam_phy))
```

Get the families and phyla that are common to the two donors

```{r}
spc_fam_both = find_don_shared(spc_don, 'Family')
spc_phy_both = find_don_shared(spc_don, 'Phylum')
spc_fam_both = spc_fam_both[spc_fam_both %in% spc_fam_phy$Family]
spc_phy_both = spc_phy_both[spc_phy_both %in% spc_fam_phy$Phylum]
```


Get database sizes for species at the family and phylum levels

```{r}
spc_fam_size = (spc_don
             %>% filter(Family %in% spc_fam_both)
             %>% count(Donor, Family, name = 'size'))
spc_phy_size = (spc_don
             %>% filter(Phylum %in% spc_phy_both)
             %>% count(Donor, Phylum, name = 'size'))
```

## Include source donor in all the coverage/recipient data frames

### Bins

```{r}
bin_cov = (bin_cov
           %>% mutate(Donor = gsub('Don', 'Donor', substr(feature, 1, 4))))
mag_cov = (mag_cov
           %>% mutate(Donor = gsub('Don', 'Donor', substr(feature, 1, 4))))
spc_rec = (spc_rec
           %>% left_join((spc_don
                          %>% select(Strain, Donor)),
                         by = 'Strain')
           %>% filter(!is.na(Donor)))
```


## Get Counts

### Bins

```{r}
bin_cov = (bin_cov
           %>% filter(coverage > 75)
           %>% left_join(bin_tax, by = c('feature' = 'BinLab'))
           %>% rename(BinLab = feature))

fbin_count = get_counts(bin_cov, bin_fam_both, 'Family', 'Bin', bin_fam_size)
pbin_count = get_counts(bin_cov, bin_phy_both, 'Phylum', 'Bin', bin_phy_size)
```

### MAGs

```{r}
mag_cov = (mag_cov
           %>% filter(coverage > 75)
           %>% left_join(bin_tax, by = c('feature' = 'BinLab'))
           %>% rename(BinLab = feature))
fmag_count = get_counts(mag_cov, mag_fam_both, 'Family', 'MAG', mag_fam_size)
pmag_count = get_counts(mag_cov, mag_phy_both, 'Phylum', 'MAG', mag_phy_size)
```

### Species

```{r}
spc_rec = (spc_rec
           %>% filter(coverage > 0.1)
           %>% select(Sample, Family, Phylum, Strain, Donor))
fspc_count = get_counts(spc_rec, spc_fam_both, 'Family', 'Species', spc_fam_size)
pspc_count = get_counts(spc_rec, spc_phy_both, 'Phylum', 'Species', spc_phy_size)
```

## Merge everything into a single data frame

```{r}
all_counts = rbind(fbin_count, pbin_count,
                   fmag_count, pmag_count,
                   fspc_count, pspc_count)

all_counts = (all_counts
              %>% left_join(dat_map, by = 'Sample')
             %>% filter(!is.na(Group))
             %>% mutate(order = case_when(Group == 'ABAB' ~ 'mixed',
                                          (Group == 'AB' & Donor == 'DonorA') |
                                            (Group == 'BA' & Donor == 'DonorB') ~
                                            'first',
                                          TRUE ~ 'second'),
                        order = factor(order,
                                       levels = c('mixed', 'second', 'first'))))
```

## Modeling

### Run the models on everything

Run each feature level through the models

```{r}
labs = expand.grid(c('f','p'), c('bin','mag', 'spc'))
labs = paste(labs[,1], labs[,2], sep = '')
sets = expand.grid(TaxLev = c('Family', 'Phylum'),
                   FeatureType = c('Bin', 'MAG', 'Species'))
sets$labs = labs

all_out = list()
for (i in 1:nrow(sets)){
  all_out[[sets$labs[i]]] = use_loop_mods(all_counts,ft = sets$FeatureType[i],
                                             tl = sets$TaxLev[i])
}
```

### Diagnostics

```{r}
prdfl = lapply(all_out, function(x) x[['prdf']])
all_prdf = (do.call(rbind, prdfl)
            %>% mutate(TaxLev = factor(TaxLev, levels = c('Phylum', 'Family')),
                       feature = factor(feature),
                       feature = order_levs(feature, TaxLev)))
ggplot(all_prdf, aes(fitted, resid)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(method = 'lm') +
  facet_wrap(~feature, scales = 'free')
ggplot(all_prdf, aes(fitted, abs(resid))) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(method = 'lm') +
  facet_wrap(~feature, scales = 'free')
ggplot(all_prdf, aes(sample = resid)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~feature, scales = 'free')
```

Some of the phylum models are okay, but all of the family models are terrible. I
can't really say that I would trust any of these outputs.

### Results

Get the contrast estimates and p-values

```{r}
all_mods = lapply(all_out, function(x) x[['mods']])
all_contr_pv = lapply(all_out, function(x) get_contr_pv(x[['mods']]))
for (lab in names(all_contr_pv)){
  all_contr_pv[[lab]] = mutate(all_contr_pv[[lab]],
                               FeatureType = sets$FeatureType[which(sets$labs == lab)],
                               TaxLev = sets$TaxLev[which(sets$labs == lab)])
}
```

Make sure the full family and phylum information is present for all the tests

```{r}
all_fam_phy = rbind(mutate(bin_fam_phy, FeatureType = 'Bin'),
                    mutate(bin_fam_phy, FeatureType = 'MAG'),
                    mutate(spc_fam_phy, FeatureType = 'Species'))
all_contr_df = do.call(rbind, all_contr_pv)

all_contr_df = (all_contr_df
                %>% rename(Family = feature)
                %>% left_join(all_fam_phy, by = c('FeatureType', 'Family'))
                %>% mutate(Phylum = case_when(TaxLev == 'Phylum' ~ Family,
                                              TRUE ~ Phylum),
                           Family = case_when(TaxLev == 'Phylum' ~ 'Full Phylum',
                                              TRUE ~ Family),
                           Family = relevel(factor(Family), 'Full Phylum'))
                %>% select(-n))
```

#### Visualization

There are a few different ways to visualize this. The first is like we did with the features:

```{r}
ggplot(all_contr_df, aes(Family, estimate, colour = within_don)) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_linerange(aes(ymin = estimate - (SE*1.96),
                 ymax = estimate + (SE*1.96)),
                 position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = 2) +
  coord_flip() +
  scale_x_discrete(limits = rev) +
  facet_grid(Phylum ~ FeatureType + numerator, scales = 'free') +
  scale_colour_manual(values = c("DonorA" = "#7b3294",
                                  "DonorB" = "#008837"),
                     name = 'Donor') 
  
```

A second option is to look at all the families vs all the phyla

```{r}
ggplot(all_contr_df, aes(estimate, colour = TaxLev)) +
  geom_density() +
  facet_grid(numerator ~ FeatureType) +
  scale_colour_brewer(palette = 'Set1')
```

If the effect is bigger within family than within phylum, we expect the phylum
curves to be narrower than the family curves. I guess we can sort of see that if
we squint, but it's very messy. The problem really is that there aren't enough
families that are present in both donors to make for a big enough dataset to
test this.